---
title: "South American Transition Zone Poaceae"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

This is the code to obtain and process the data for the assessment of niche shifts and Cenocron assignation of MTZ's plants.

# Set up

## Libraries

```{r}
library(tidyverse)
library(ape)
library(ggtree)
library(deeptime)

library(rgbif)
library(CoordinateCleaner)

library(sf)
library(tmap)
library(kewr)
sf_use_s2(F)

data("World")
source('C:/Users/USER/Desktop/R projects/theme_virix.r')
```

## Phylogeny

Phylogeny obtained from ["Plant and Fungal Trees of Life Project (PAFTOL) at the Royal Botanic Gardens"](http://treeoflife.kew.org "PAFTOL 2025").

First we remove all the tips that are not species.

```{r}
tree_all <- read.tree(file = "data/trees/treeoflife.3.0.tree" )
  
# Seleccionar las asteráceas

all_sp <- tree_all$tip.label

to_prune <- str_subset(all_sp, 'Poales_Poaceae_', negate = T) %>%
  # Quitar puntas  no identificadas a especie 
  c(., str_subset(tree_all$tip.label, '_sp.'))

tree1 <- drop.tip(tree_all, to_prune)

# Quitar duplicados
# Get species name by removing the last underscore and everything after it
species_names <- sub("_[^_]+$", "", tree1$tip.label)

# Keep only the first occurrence of each species
tips_to_drop <- tree1$tip.label[duplicated(species_names)]
clean_tree <- drop.tip(tree1, tips_to_drop)

# simplify names
clean_tree$tip.label <- str_remove(clean_tree$tip.label,
                                   'Poales_Poaceae_') %>%
  word(1,2, sep = '_')

```


## Geographical setting

The Mexican Transition Zone sensu Escalante et al.

```{r}
satz_sf <- st_read('data/SATZ.gpkg') 

qtm(satz_sf)

wkt_polygon<- st_as_text(satz_sf$geom)  %>% wkt_parse()
```

# Get GBIF ids to standardize taxonomi

Obtain the unique ID number in GBIF of each tree tip, as well as the presence of it in the MTZ.

```{r}
source('occ_explore.R')
library(RPushbullet)

all_sp <- clean_tree$tip.label 

lapply(all_sp, gbif_keys) 
 
 
multiple_matches <- pull(read.csv('not found.csv', header = T), label) %>%
  unique()

lapply(multiple_matches, gbif_keys_synonims)

pbPost(type = 'note',title = 'GBIF Keys Stopped',
       body = paste('Avance de ',1  
                    scales::percent(
                    length(checked)/length(all_sp))
                    ))


```

## Read the final database

```{r}
gbif_keys_all <- read.csv('gbif_keys.csv') %>%
  filter(occ> 0)%>% 
  mutate(genus = word(label, 1, sep = '_')) %>%
  distinct()


gbif_keys_all %>%
  ggplot(aes(x = occ)) + 
  geom_histogram(binwidth = 10, fill = 'darkolivegreen' ) +
  theme_virix() +
  labs(x = 'Número de registros',
       y = 'Especies',
       title = 'Registros por especie de Poáceas en la SATZ',
       subtitle = 'Conteo preeliminar')+
  coord_cartesian(expand = F)

```


## Summarize

```{r}
gbif_keys_all 
n_genus <- gbif_keys_all %>% group_by(genus) %>% tally() %>% nrow()
n_sp <- gbif_keys_all %>% group_by(label) %>% tally() %>% nrow()
n_occ_sp <- gbif_keys_all %>% pull(occ) %>% mean() %>% round(., 0)

data_frame(level = c('Genus', 'Species', 'Mean occ per species'),
           number = c(n_genus, n_sp, n_occ_sp))
           
```

# Occurrence search


## Keys to look up

```{r}
keys <- gbif_keys_all %>% 
  pull(usagekey) 

length(keys)
```

## Occurrence download

```{r}
gbif_download <- occ_download(
    type="and",
    #Taxon Keys
    pred_in("taxonKey", keys),
    # With coordinates and no issues
    pred("hasCoordinate", TRUE),
    pred("hasGeospatialIssue", FALSE),
   #pred_within('geometry', wkt_polygon),
    #Year
    #pred_gte("year", 1950),  
    # Basis of record
    #pred_in("basisOfRecord",'PRESERVED_SPECIMEN'),
    # Only Native species
   pred_or(pred_not(
     pred("establishmentMeans","MANAGED")),
      pred_not(pred_notnull("establishmentMeans"))),
   pred_or(pred_not(pred("establishmentMeans","INTRODUCED")),
           pred_not(pred_notnull("establishmentMeans"))),
   pred_or(pred_not(pred("establishmentMeans","INVASIVE")),
           pred_not(pred_notnull("establishmentMeans"))),
   pred_or(pred_not(pred("establishmentMeans","NATURALISED")),
           pred_not(pred_notnull("establishmentMeans"))),
    # Only records with low uncertanty
    pred_or(pred_lt("coordinateUncertaintyInMeters",1000),
            pred_not(pred_notnull("coordinateUncertaintyInMeters"))),
    format = "SIMPLE_CSV"
    ) 

occ_download_wait(gbif_download)

```


```{r}


gbif_result <-  occ_download_get(key = '0000833-250515123054153') %>%
  occ_download_import()

```

# Data cleaning
## Remove species with less 

```{r echo=FALSE, results= 'asis'}
paste('The download has', nrow(gbif_result ), 'records', 
      n_distinct(gbif_result$species), 'species,',
      n_distinct(gbif_result$genus), 'genus, from',
      n_distinct(gbif_result$countryCode), 'countries.') %>% cat()
```

First, we clean using `CoordinateCleaner` to remove invalid coordinates, points at sea, at institutions, duplicates, and with equal lat-lon values.

```{r}
gbif_result_clean <- gbif_result %>% 
  select(occurrenceID, speciesKey, order, family, species,
         decimalLatitude, decimalLongitude, establishmentMeans) %>%
    cc_val(lat = 'decimalLatitude', lon = 'decimalLongitude') %>%
    cc_sea(lat = 'decimalLatitude', lon = 'decimalLongitude') %>%
    cc_equ(lat = 'decimalLatitude', lon = 'decimalLongitude') %>%
    cc_inst(lat = 'decimalLatitude', lon = 'decimalLongitude') %>%
  cc_zero(lat = 'decimalLatitude', lon = 'decimalLongitude') %>%
  cc_dupl(lat = 'decimalLatitude', lon = 'decimalLongitude')
```

## Turn into spatial object

Then, the database is turned into an spatial object to count how many points per species are inside the satz.

```{r}
gbif_sf <- gbif_result_clean   %>%
  mutate(lon = decimalLongitude,
         lat = decimalLatitude) %>%
  rename(usagekey = speciesKey) %>% 
  filter(!establishmentMeans %in% c('Introduced','Uncertain', 'Vagrant', 'NativeReintroduced')) %>% 
  left_join(.,gbif_keys_all[c(1,7)], by= 'usagekey') %>%
  st_as_sf(coords = c('decimalLongitude',
                      'decimalLatitude' ), 
           crs = st_crs(satz_sf)) %>%
    
  st_join(satz_sf) %>% 
  mutate(in_satz = if_else(is.na(Region), 0, 1)) %>%
  rownames_to_column('ID') %>%
  group_by(species) %>%
  # Add record count
  mutate(n = n_distinct(ID),
            satz = sum(in_satz,na.rm = T),
            perc_zt = satz/n,
            # Calcular los cortes solo una vez fuera del mutate
             endem_class = cut(perc_zt,
                               breaks = c(0,0.05, 0.1,0.5, 0.99 ,1),
                               include.lowest = TRUE,
                               right = FALSE))  


```



## Count species endemic to the satz

```{r}
gbif_sf %>%
  #filter(perc_zt > 0.005) %>%
  group_by(label, endem_class) %>%
  reframe(perc_zt = mean(perc_zt)) %>%
  ggplot(aes(x = perc_zt, fill = endem_class)) +
  geom_histogram(binwidth = 0.01) +
  theme_virix() +
  labs(x = 'Porcentaje de Registros en SATZ',
       y = 'Número de especies',
       fill = 'Endemismo',
       title = 'Endemismos de Poáceas en la SATZ',
       subtitle = '¿Eliminar <0.01 registros en SATZ?')+
  coord_cartesian(expand = F) +
  theme(legend.position = 'right')
```
```{r}
gbif_sf %>%
  group_by(label, endem_class) %>% 
  summarise(perc_zt = mean(perc_zt)) %>%
  ggplot(aes(x = perc_zt, col = perc_zt)) +
  geom_sf(size = 0.1, alpha = 0.1) + 
  facet_wrap(endem_class~.) +
  scale_color_viridis_c() +
  theme_virix() + theme(legend.position = 'none')
```




```{r echo=FALSE,  fig.asp= 0.5, dpi=150}
data('World')

n_occ_sp <- gbif_sf %>% 
  filter(perc_zt >0.05) %>%
  nrow() %>% scales::number()

credits <-paste('The database has',
                n_occ_sp, 'records and',
                scales::number(n_distinct(gbif_sf$label)), 'species.')


tm_shape(World, bbox = st_bbox(gbif_sf)) +
  tm_fill('gray90')+
tm_shape(filter(gbif_sf, perc_zt >0.05)) +
  tm_dots('perc_zt',fill_alpha = 0.2,
          fill.scale = tm_scale_continuous(values = '-viridis'))  +
  tm_title(str_wrap(credits, 35)) + tm_layout(legend.position = c('left','bottom'))
```


# Save everything

## Prune tree
```{r}
sp_list_final <- unique(gbif_sf$label) 

# Remove tips with no records 
final_tree <- drop.tip(clean_tree, 
                   clean_tree$tip.label[!clean_tree$tip.label %in% 
sp_list_final])

final_tree

write.nexus(final_tree, file = 'data/final_tree_poaceae.nex') 

```


# Save occurrence data 
```{r}
st_drop_geometry(gbif_sf) %>%
  filter(perc_zt >0.05) %>%
  write.csv('data/occurrences_poaceae.csv', row.names = F)
```

# Plotting

```{r, fig.asp= 1}

library(ggrepel)

tr_dat<- gbif_sf %>%
  st_drop_geometry() %>%
  group_by(label, endem_class, perc_zt) %>% 
  tally() %>% na.omit() 

tr_dat


```

```{r, fig.asp= 0.7, fig.width= 8, dpi=600}

ggtree(final_tree, ladderize = T,layout = 'circular', lwd = 0.2) %<+% tr_dat  + 
  geom_tiplab(size = 1 )  + 
  geom_tippoint(aes(col = perc_zt), size = 1) + theme_void() +
  scale_color_viridis_c(direction = -1) +
  labs(color = 'Proporción de\nregistros en SATZ')


```

